Chemoinformatics: The Computer Science of Chemistry

### 2011-06-22 paper.tex
reworked the open source and open data section

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  technologies to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline -- chemoinformatics, the use of computers to discover new molecules with
  desired behavior, such as new pharmaceuticals or industrial solvents. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new "miracle molecule". Only within the last decade -- and largely because of
  chemists volunteering their time for an Open Science "movement" -- do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront hundreds of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

@@

The motivation for this article is to highlight a variety of topics in
cheminformatics and present the computational and algorithmic problems that are
addressed i the field. The hope is that we will be able to highlight areas where modern CS
research could contribute to these problems.

@@

See e.g. role for the EBI and genome campus,
https://registration.hinxton.wellcome.ac.uk/display_info.asp?id=246 See also
impact of open chemical data, ChEMBL to future, e.g. ELIXIR
http://www.elixir-europe.org/page.php

@@

A key success of the open source movement is that experts can team up for
collaborating in their expert areas. Especially in the chemistry domain, with a
strong expert system history, this is even more relevant due to its complex data
types and relation to many outcome variables, like biological activities,
toxicological effects, influence to diseases, or effects on the environment.

Open Babel and the Chemistry Development Kit (CDK) are two of the most widely
used Open Source chemical toolkits. Such toolkits are used, for example, to
read/write chemical file formats, manipulate chemical structures, measure
similarity of molecules, search for substructures, and generate 2D diagrams. The
underlying algorithms used include graph algorithms (e.g. maximal common
substructure, canonical labelling of graphs), geometrical methods (e.g. Kabsch
alignment) and vector manipulation (e.g. converting coordinates in various
systems, 3D structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have been
developed that rely on these toolkits to handle chemical data; for example, the
molecular viewer Avogadro (http://avogadro.openmolecules.net) uses Open Babel,
while the molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

Open Babel (http://openbabel.org) is composed of a set of command-line
applications, a GUI, and a C++ library with SWIG-generated bindings in several
languages. It runs cross-platform and uses CMake and CTest for building and
testing. One of its unique features is support for a large number of chemical
file formats and its babel command-line application is widely used for
conversion between them. The origin of Open Babel is OELib, a chemical toolkit
released under the GPL by OpenEye Scientific Software. In 2001, OpenEye decided
to discontinue work on OELib and to focus instead on an internal rewrite of the
library which would not be open source (OEChem). At that point, OELib was
renamed to Open Babel and since then it has been maintained and developed by a
community led by Dr Geoffrey Hutchison (University of Pittsburgh).

The Chemistry Development Kit (http://cdk.sf.net, \cite{Steinbeck2006}) project
originated in 2000 as a common base for two existing Java applications Jmol and
JChemPaint. The initial code bases inherited from both these project and
CompChem. It soon outgrew this initial goal and is now a fully-featured
cheminformatics toolkit in its own right. The CDK is a Java library and uses Ant
and JUnit for building and testing, as well as PMD and OpenJavaDocCheck for code
quality control. A particular strength of the toolkit is its support for a large
number of so-called “molecular descriptors” (numerical representations of
chemical structures used in machine learning to link chemical structure to a
physical or biological property). The project was initially led by Dr Christoph
Steinbeck (EMBL-EBI, UK) and Dr Egon Willighagen (Karolinska Institutet,
Sweden), and more recently Dr Rajarshi Guha (NIH, USA) joined as third, current
project leader.

Both the CDK and Open Babel are very open to new contributors to the projects.
They both have an open mailing list, without any further private lists.
Likewise, issue tracking systems are fully open, and ongoing and future
development is openly discussed. Contributors are encouraged to get involved
with any of the following areas: development of new features, improvement of
existing features, performance optimisation, fixing bugs, documentation, website
development, reporting bugs, and helping users on the mailing list. Anyone
interested should email the respective developer’s mailing list (either
cdk-devel@lists.sf.net or openbabel-devel@lists.sf.net) with their interests and
background, and a mentor will be assigned to help them join the project.

Contributions as such are made to both projects on a regular basis, both from
industries more interested in the use and application of these libraries, as
well as from research groups in academia. And example here is the contribution
of a ring search algorithm replacing the earlier Figueras algorithm by
mathematicians from the TU M\"{u}nchen in Germand (\cite{Berger2004}). The former is
maintained in the CDK, as the project also aims to provide an educational
resource.

Challenges faced by the CDK and OpenBabel libraries are abundant. Many of these
originate from the complexity of the graph theory used in chemistry, which is an
approximate but insufficient representation to capture chemical reality.
Existing areas of interest are enumeration of colored graphs, taking into
account symmetry, and symmetry detection itself, which not merely derives from
the chemical graph, but typically also from the 3D geometry of the chemical.
However, more challenging is that chemical representation of a molecule really
needs to capture non-static graphs, in order to take into account delocalisation
and tautomeric phenomena molecules can practically undergo.

>>>>

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline -- chemoinformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  including the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new "miracle molecule". Only within the last decade -- and largely because of
  chemists volunteering their time for an Open Science "movement" -- do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety
  of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

@@

The motivation for this article is to highlight a variety of topics in
cheminformatics and present the computational and algorithmic problems
that are addressed in the field. The hope is that we will be able to
highlight areas where modern CS research could contribute to these
problems.

@@

A key aspect of cheminformatics is that much of its development
requries access to collections of molecular structures (and usually,
associated bioactivity data). As has been noted, it is only recently
that large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale benchmarking of
algorithms. The latter is especially relevant for the plethora of
data mining techniques that are emoloyed in cheminformatics. At this
point, the nature of Open Data focuses primarily on structure and
actvity data types. On the other hand, there is a distinct lack of
textual data (such as journal articles) that are Open in nature. While
PubMed abstract serve as a proxy for journal articles, text mining
methods in cheminformatics are hindered by this state of affairs.
Interestingly, patent information is publicly accessible (cf. Google
Patents) and there have been efforts to mine these for chemical
information using Open tools.

Obviously freely accessible data is important for research; but
equally important are the tools to work with chemical data -- more
specifically, chemical structures. Chemical toolkits are the workhorses
of cheminformatics research and a variety of toolkits are available,
both commericial (which, in many cases, provide free academic
licenses) and Open Source. In this article we do not debate the
arguments for and against Open Source; rather, we focus on Open Source
tools since this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

@@

A number of toolkits are available including the Chemistry Development
Kit (CDK), OpenBabel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(http://avogadro.openmolecules.net) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
OpenBabel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.

While algorithmic challenges are certainly faced by toolkits, software
engineering challenges probably affect the use and development of the
toolkits on a daily basis. Examples include verifying code coverage,
thread safety, documentation quality and so on. Some projects such as
the CDK employ a variety of tools to address some of these issues.
Examples include the use of PMD to analyze code quality and JUnit to
perform unit testing. At the level of implementation, there are still
inefficiencies in some toolkits. For example the CDK employs a JJTree
parser to implement the SMARTS specification, but employs a hand built
parser to handle SMILES. The latter is essentially a complex tangle of
\emph{if \ldots then} statements and is not easily maintainable.

It is also important to note that certain challenges faced by nearly
all cheminformatics toolkit stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality -- and edges cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. However, more challenging is that chemical
representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and tautomeric
phenomena molecules can practically undergo. It is important to
realize that though chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems.

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and OpenBabel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimisation, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developer’s mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.
